Compositional equivalences based on Open pNets

### 2021-10-06 2016-Weak-FH-Bisim/WeakBisim.tex
spec and impl in appendix

The disjoint union on sets is $\uplus$. We extend it to  disjoint union  of indexed 
sets defined by the merge of the 
two sets provided they are indexed on disjoint families.
The elements
of the union of two indexed sets are then accessed by using an index of one of the two
joined families.
%We suppose here that disjoint unions are always well defined; this
%requires to choose disjoint  
%indices in the definition of the component system (e.g. name of
%methods of different interfaces).
$\setminus$ is the standard subtraction operation on indexed sets: $\dom(A\setminus B)=\dom(A)\setminus B$.

@@

This article also uses substitutions. Applying a substitution inside a term $t$ is denoted $t\subst{y_i\gets e_i}^{i\in I}$ and consists in replacing in parallel all the occurrences of variables $y_i$ in the term $t$ by the terms $e_i$. Note that a substitution is defined by a partial function that is applied on the variables inside a term. We let $\Post$ range over partial functions that are used as substitution and use the notation $\{y_i\gets e_i\}^{i\in I}$ to define such a partial function\footnote{When using this notation, we suppose, without loss of generality that each $y_i$ is different.}. These partial functions are sometimes called \emph{substitution functions} in the following.
 $\subst{\Post}$ is thus the operation that applies,  in a parallel manner,  the substitution defined by the partial function $\Post$. $\otimes$ is a composition operator on these partial functions, such that for any term $t$ we have: $t\subst{\Post\shortotimes\Post'} = (t\subst{\Post'})\subst{\Post}$.
This property must also be valid when the substitution does not operate on all variables.
We thus define a composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\

@@

A pNet $Q$ is \emph{closed} if it has no hole: $\Holes(Q)=\emptyset$; else it
is said to be \emph{open}.
Sort comes naturally with a compatibility relation that is similar to a type-compatibility check. We simply say that two sorts are compatible if they consist of the same actions with the same arity. In practice, it is sufficient to check the equality of the two sets of pairs $(\alpha,n)$ of the two sorts\footnote{A more complex compatibility relation could be defined, but this is out of the scope of this article.}.
\end{definition}

@@

We define now a bisimulation relation tailored to open automata and their parametric nature. This relation relates states of the open automaton and guarantees that the related states are observationally equivalent, i.e. equivalent states can trigger transitions with identical action labels. Its key characteristics are 1) the introduction of predicates in the bisimulation relation: as states may contain variables, relation between states may depend on the value of the variables; 2) the bisimulation property relates elements of the open transitions and takes into account predicates over variables, actions of the holes, and state modifications.
 We name it FH-bisimulation,
 as a short cut for the ``Formal Hypotheses'' over the holes behaviour manipulated in the
 transitions, but also as a reference to the work of De Simone~\cite{deSimone85},
 that pioneered this idea.

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{XFIG/SPSpecOpen}}
   \caption{Open Automaton of  $\symb{SimpleProtocolSpec}$}
   \label{SimpleProtCounter:SpecOA}
\end{figure}

@@

 \begin{figure}[ht]
  \centerline{\includegraphics[width=15cm]{XFIG/SPImplOpen}}
  \caption{Open Automaton of $\symb{SimpleProtocolImpl}$}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

@@

\subsubsection*{Running example}
\label{subsubsection:runnig example}
In Section~\ref{sec:weak} we have shown the full saturated weak automaton for both \symb{SimpleProtocolSpec} and \symb{SimpleProtocolImpl}. We will show here how we can check if some given relation between these two automata is a weak FH-Bisimulation.

@@

@@

\label{Appendix:FullExample}

@@

The first piece of code is the textual definition of the \TODO{SimpleSystemSpec}
pNet (named \TODO{SimpleProtSpec2} in the code here), that was drawn in Figure \ref{SimpleProt:Impl}, page \pageref{SimpleProt:Impl}. This code should be intuitive enough to read, with the following
language conventions, that brings some user-friendly features, mapped by the editor into pure pNet constructs.

@@

\begin{lstlisting}[basicstyle=\scriptsize\ttfamily, language=java, frame=single]
SimpleProtSpec2:
import "Data_Alg.algp"
root SimpleProtSpec2
const in, out:Action
const p_send, q_recv: Action
const tau:Action

@@

pNet SimpleProtSpec2
holes P,Q
subnets P,PerfectBuffer,Q
vars p_a,q_b:Action m:Data ec:Nat

@@

Next is the code for the \TODO{SimpleSystem implementation} pNet (named \TODO{SimpleProtImpl2}):
\medskip
\begin{lstlisting}[basicstyle=\scriptsize\ttfamily, language=java, frame=single]
SimpleProtImpl2:
  import "Data_Alg.algp"
  root SimpleProtImpl2
const in,out:Action
const tau,p_send,q_recv,m_recv,m_send,m_error: Action
const s_recv,s_send,s_ack,s_error,r_recv,r_ack,r_send: Action

@@

pNet SimpleProtImpl2
  holes P,Q
  subnets P,SimpleProtocol,Q
  vars p_a,q_a:Action  m:Data c:Nat
vector SV0 <p_send(m),in(m),_>->synchro(in(m))
vector SV1 <p_a,_,_>->p_a [p_a != p_send(x)]
vector SV2 <_,out(m,ec),q_recv(m,ec)>->synchro(out(m,ec))
vector SV3 <_,_,q_b>->q_b [q_b != q_recv(x,y)]
\end{lstlisting}

@@

\begin{figure}[h]
   \centerline{\includegraphics[width=10cm]{XFIG/SimpleProtImpl-WOA2}}
   \caption{Weak Open Automaton of the \TODO{implementation}}
   \label{Appendix:ImplOA2}
 \end{figure}

In Figure \ref{Appendix:ImplOA2} we recall the weak open automaton of our implementation pNet.

This drawing is based on the observation that states 202 and 000 are only linked by a "pure $\tau$" transition, and have exactly the same possible behaviours.
%More tricky, they do not have the same state variables, as 202 has the variables of state $s2$, but none of these variables are used in transitions from 202.
In this configuration we can guarantee that they are Weak bisimilar, and we have merged their (incoming and outgoing) transitions in the figure. We denote this 
equivalence class of states as $\{000,202\}$.

@@

         Note: if we were using the alternative Weak bisimulation relation from Appendix \ref{app-WFH-equiv}, Lemma \ref{lem-rel-OT-WOT}, that is checking strong bisimulation between the corresponding 
weak automaton, we would have a more transitions coverage to examine, as we have 4 weak transitions for $b0$ in the \TODO{specification} weak automaton, and 7 WOTs (including 4 
parameterised WOTs) from 000 in the \TODO{implementation} automaton.

Preliminary remarks:
  \begin{itemize}
    \item Both pNets trivialy verify the ``non-observability''
      condition: the only vectors having $\tau$ as an action of a
      sub-net are of the form ``$< -, \tau, -> -> \tau$''.
    \item We must take care of variable name conflicts: in our example, the variables of the 2 systems already have different names, but the action parameters occurring in the 
transitions (m, msg, ec) are the same, that is not correct. In the tools, this is managed by the static semantic layer; in the following example, we have renamed all conflicting 
variables using suffix 1 for the \TODO{Specification}, and 2 for the \TODO{Implementation}.
    \end{itemize}

>>>>

The disjoint union on sets is $\uplus$. We extend it to  disjoint union  of indexed 
sets defined by the merge of the 
two sets provided they are indexed on disjoint families.
The elements
of the union of two indexed sets are then accessed by using an index of one of the two
joined families.
%We suppose here that disjoint unions are always well defined; this
%requires to choose disjoint  
%indices in the definition of the component system (e.g. name of
%methods of different interfaces).
The standard subtraction operation on indexed sets is $\setminus$, with $\dom(A\setminus B)=\dom(A)\setminus B$.

@@

This article also uses substitutions. Applying a substitution inside a term $t$ is denoted $t\subst{y_i\gets e_i}^{i\in I}$ and consists in replacing in parallel all the occurrences of variables $y_i$ in the term $t$ by the terms $e_i$. Note that a substitution is defined by a partial function that is applied on the variables inside a term. We let $\Post$ range over partial functions that are used as substitution and use the notation $\{y_i\gets e_i\}^{i\in I}$ to define such a partial function\footnote{When using this notation, we suppose, without loss of generality that each $y_i$ is different.}. These partial functions are sometimes called \emph{substitution functions} in the following. Thus,
 $\subst{\Post}$ is  the operation that applies,  in a parallel manner,  the substitution defined by the partial function $\Post$. $\otimes$ is a composition operator on these partial functions, such that for any term $t$ we have: $t\subst{\Post\shortotimes\Post'} = (t\subst{\Post'})\subst{\Post}$.
This property must also be valid when the substitution does not operate on all variables.
We thus define a composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\

@@

A pNet $Q$ is \emph{closed} if it has no hole: $\Holes(Q)=\emptyset$; else it
is said to be \emph{open}.
Sort comes naturally with a compatibility relation that is similar to a type-compatibility check. We simply say that two sorts are compatible if they consist of the same actions with the same arity. In practice, it is sufficient to check the equality of the two sets of action signatures of the two sorts\footnote{A more complex compatibility relation could be defined, but this is out of the scope of this article.}.
\end{definition}

@@

We define now a bisimulation relation tailored to open automata and their parametric nature. This relation relates states of the open automaton and guarantees that the related states are observationally equivalent, i.e. equivalent states can trigger transitions with identical action labels. Its key characteristics are 1) the introduction of predicates in the bisimulation relation: the relation between states may depend on the value of the variables; 2) the bisimulation property relates elements of the open transitions and takes into account predicates over variables, actions of the holes, and state modifications.
 We name it FH-bisimulation,
 as a short cut for the ``Formal Hypotheses'' over the holes behaviour manipulated in the
 transitions, but also as a reference to the work of De Simone~\cite{deSimone85},
 that pioneered this idea.

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{XFIG/SPSpecOpen}}
   \caption{Open automaton for  $\symb{SimpleProtocolSpec}$}
   \label{SimpleProtCounter:SpecOA}
\end{figure}

@@

 \begin{figure}[ht]
  \centerline{\includegraphics[width=15cm]{XFIG/SPImplOpen}}
  \caption{Open automaton for $\symb{SimpleProtocolImpl}$}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

 \TODO{a counter example in  CCS}

@@

\TODO{Conclude with CCS as an example}
\subsubsection*{Running example}
\label{subsubsection:runnig example}
In Section~\ref{sec:weak} we have shown the full saturated weak automaton for both \symb{SimpleProtocolSpec} and \symb{SimpleProtocolImpl}. We will show here how we can check if some given relation between these two automata is a weak FH-Bisimulation.

@@

\TODO{LUDO idees: \cite{Beohar:2020} conditions de transitions et bisim basees la dessus, leur bisimulation est plus logique et plus riche de ce cote, la notre plus structurelle et plus orientee composition, de facon liee ils voient l environnement comme repondant a des relaations l=ogiques avec certaines rpoprietes, on ne prend pas en compte un environnement mais on a des regles de compositions entre systemes et etablit une bisim prenant ca en compte}

@@

\label{Appendix:FullExample}
\TODO{Ludo: spec implem dans l annexe}

@@

The first piece of code is the textual definition of the $\symb{SimpleProtocolSpec}$
pNet, that was drawn in Figure \ref{SimpleProt:Impl}, page \pageref{SimpleProt:Impl}. This code should be intuitive enough to read, with the following
language conventions, that brings some user-friendly features, mapped by the editor into pure pNet constructs.

@@

\begin{lstlisting}[basicstyle=\scriptsize\ttfamily, language=java, frame=single]
SimpleProtocolSpec:
import "Data_Alg.algp"
root SimpleProtocolSpec
const in, out:Action
const p_send, q_recv: Action
const tau:Action

@@

pNet SimpleProtocolSpec
holes P,Q
subnets P,PerfectBuffer,Q
vars p_a,q_b:Action m:Data ec:Nat

@@

Next is the code for the \symb{SimpleProtocolImpl} pNet:
\medskip
\begin{lstlisting}[basicstyle=\scriptsize\ttfamily, language=java, frame=single]
SimpleProtocolImpl:
  import "Data_Alg.algp"
  root SimpleProtocolImpl
const in,out:Action
const tau,p_send,q_recv,m_recv,m_send,m_error: Action
const s_recv,s_send,s_ack,s_error,r_recv,r_ack,r_send: Action

@@

pNet SimpleProtocolImpl
  holes P,Q
  subnets P,SimpleProtocol,Q
  vars p_a,q_a:Action  m:Data c:Nat
vector SV0 <p_send(m),in(m),_>->synchro(in(m))
vector SV1 <p_a,_,_>->p_a [p_a != p_send(x)]
vector SV2 <_,out(m,ec),q_recv(m,ec)>->synchro(out(m,ec))
vector SV3 <_,_,q_b>->q_b [q_b != q_recv(x,y)]
\end{lstlisting}

@@

\begin{figure}[h]
   \centerline{\includegraphics[width=10cm]{XFIG/SimpleProtImpl-WOA2}}
   \caption{Weak Open Automaton for \symb{SimpleProtocolImpl}}
   \label{Appendix:ImplOA2}
 \end{figure}

In Figure \ref{Appendix:ImplOA2} we recall the weak open automaton of \symb{SimpleProtocolImpl}. 
This drawing is based on the observation that states 202 and 000 are only linked by a "pure $\tau$" transition, and have exactly the same possible behaviours.
%More tricky, they do not have the same state variables, as 202 has the variables of state $s2$, but none of these variables are used in transitions from 202.
In this configuration we can guarantee that they are Weak bisimilar, and we have merged their (incoming and outgoing) transitions in the figure. We denote this 
equivalence class of states as $\{000,202\}$.

@@

         Note: if we were using the alternative Weak bisimulation relation from Appendix \ref{app-WFH-equiv}, Lemma \ref{lem-rel-OT-WOT}, that is checking strong bisimulation between the corresponding 
weak automaton, we would have a more transitions coverage to examine, as we have 4 weak transitions for $b0$ in the \symb{SimpleProtocolSpec} weak automaton, and 7 WOTs (including 4 
parameterised WOTs) from 000 in the \symb{SimpleProtocolImpl} automaton.

Preliminary remarks:
  \begin{itemize}
    \item Both pNets trivially verify the ``non-observability''
      condition: the only vectors having $\tau$ as an action of a
      sub-net are of the form ``$< -, \tau, -> -> \tau$''.
    \item We must take care of variable name conflicts: in our example, the variables of the 2 systems already have different names, but the action parameters occurring in the 
transitions (m, msg, ec) are the same, that is not correct. Recall that we disambiguate the reference to the variable $m$ into $m1$ for \symb{SimpleProtocolSpec} and $m2$ for \symb{SimpleProtocolImpl}.
    \end{itemize}
